Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-01-10 analysis/paper.tex
Prepare draft

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 82 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. Modelled ranges do not
reliably predict the observed occurrence of specific species at
archaeological sites, however {[}\ldots{]} The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

The Pleistocene--Holocene transition in West Asia marked a turning point
in global environmental history, as humans brought the first plants
under cultivation and began modifying surrounding ecosystems to support
their own subsistence. West Asia is part of the native range of a
remarkable number of domesticable plant species, including wild
relatives of wheat, barley, peas, lentils, and other crops of global
importance \citep{cite}. These species supported uniquely dense and
complex Late Epipalaeolithic (15--11.7 ka) societies based on intensive
foraging \citep{cite} and eventually pre-domestication cultivation
\citep{cite}. As they were brought under domestication in the
Pre-Pottery Neolithic period (11.7--8.5 ka), the world's first
agriculture was shaped by the ecosystems from which it emerged and was
embedded in.

Decades of research in archaeobotany and zooarchaeology have
reconstructed the subsistence economies of Late Epipalaeolithic and
Neolithic sites in great detail \citep{cites}. Together with studies of
charcoal \citep{cite}, pollen \citep{cite}, soil \citep{cite}, and a
variety of palaeoclimate archives \citep{JonesEtAl2019}, they also tell
us much about the environments surrounding these settlements. However,
each of these sources of evidence is subject to the wide variety of
taphonomic and recovery biases that are inherent in any direct record of
the past. They are also, by definition, records of the (human)
environment at particular times and places. Interpolating these
snapshots to give a holistic picture of the regional ecologies is not
straightforward -- to date, it has tended to rely on non-explicit,
inductive modelling. The majority are also filtered through human
action, producing a mixed single that makes it difficult to disentangle
anthropic effects from the background of environmental change in this
period of rapid climatic alteration.

@@

\begin{itemize}
\tightlist
\item
  The transition to agriculture in West Asia was\ldots{}
\end{itemize}

\subsection{Biogeography and agricultural
origins}\label{biogeography-and-agricultural-origins}

\begin{itemize}
\tightlist
\item
  Has always been important in study of agricultural origins

  \begin{itemize}
  \tightlist
  \item
    Historically: Vavilov, Pumpelly \& Childe
  \item
    Genetic studies tell us origin points, but not ranges
  \end{itemize}
\item
  Important to e.g.

  \begin{itemize}
  \tightlist
  \item
    Distinguish environmental from potentially anthropogenic change
    \citep{MartinEtAl2016, MartinEtAl2025}
  \item
    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

  \begin{itemize}
  \tightlist
  \item
    More than the ``founder crops'';
  \item
    More than food
  \item
    (In archaebot., not all intentionally collected)
  \item
    Geographically and temporally diverse
  \item
    \ldots so we model lots of species!
  \end{itemize}
\item
  Regional ecological reconstructions generally rely on the `expert
  interpolation' (or what do they call it with isoscapes?) method

  \begin{itemize}
  \tightlist
  \item
    See CSEAS (AEA-prep) presentation
  \item
    Figure: comparisons
  \end{itemize}
\end{itemize}

\subsection{Ecological niche modelling in
archaeology}\label{ecological-niche-modelling-in-archaeology}

Ecological niche modelling or species distribution modelling is widely
used by ecologists to predict the geographic range of a species based on
a set of environmental predictors \citep{FranklinMiller2009}.
Essentially, it involves combining records of where an organism has been
observed with environmental data (climate, topography, etc.) for those
locations to model the range of environmental values at which that
species -- its environmental niche. This model can then be used to
predict the range of the organism in question either in the same or a
different environment. \citet{CITE} suggests reserving the term `species
distribution modelling' for when the method is used to recover the
verifiable range of a species in a real and existing environment, and
using `ecological niche modelling' as the broader term covering
hypothetical or predictive applications -- a convention we follow here
when referring to predictive or `hindcast' models of past ranges. Within
this overarching framework, ecological niche modelling encompasses a
wide range of applications and a variety of potential environmental
predictors, modelling approaches, and methodologies, which we will not
attempt to review here.

@@

\citet{FranklinEtAl2015} review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. In an
early application to West Asia, \citet{ConollyEtAl2012} used the
occurrence of wild and domestic \emph{Bos} remains at prehistoric
archaeological sites to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales
\citep[e.g.][]{deAndresHerreroEtAl2018, YaworskyEtAl2023} and, in a West
Asian context, of foraged plant resources in the landscape around the
Neolithic sites on the Konya Plain \citep{CollinsEtAl2018}. Modelling
the spread of crops has been another significant archaeological
application \citep[e.g.][]{CremaEtAl}, though not as yet applied to West
Asia.

@@

The major practical limitation of the hindcasting approach is that it
relies on spatially explicit, high resolution palaeoenvironmental
surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most
applications, which is perhaps why only a minority of studies use it
\citep[cf.][]{YaworskyEtAl2023}. In this study, we are able to take
advantage of the increasing availability of high resolution, global
palaeoclimate data derived from simulation experiments with general
circulation models of climate
\citep{BrownEtAl2018, BrownEtAl2020, KargerEtAl2023}.

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 97
taxa (Table~\ref{tbl-results-summary}) comprising the identifiable
species observed at at least two Late Epipalaeolithic/Pre-Pottery
Neolithic sites, according to our previous study of the regional
archaeobotanical data \citep{ArranzOtaeguiRoe2023}. This was based on
dataset collated from three previously published regional
archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
\citep{ORIGINS}, and COMPAG \citep{COMPAG}. We did not attempt to
distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

@@

Predictor data was transformed to common equal-area projection and
resolution of 5 km.

@@

@@

\begin{verbatim}
Joining with `by = join_by(taxon)`
\end{verbatim}

@@

Our results for the Levant are consistent with the current understanding
of this region as developing early intensive foraging economies
\citep[the Natufian culture,][]{cite} and as a centre of origin of
agriculture \citep{cite}. Within the Levant, many species show moderate
reductions in ranges over the Pleistocene/Holocence boundary, retreating
from the Badia/transjordan region
e.g.~\ref{fig-palaeodist-barely-levant}.

Loss of the Northern Mesopotamia--Anatolia region from the predicted
ranges of crop progenitors is interesting in light of the `golden
triangle' hypothesis
\citep{Lev-YadunEtAl2000, KozlowskiAurenche2005, AbboEtAl2010}, which
puts this region at the centre of the development of agriculture.
Multiple lines of archaeological evidence have emerged that point away
from this hypothesis and towards a more geographically diverse origin
\citep{Asouti2006, cites}, but it remains the area with some of the
earliest clear evidence of domestication
\citep{ZoharyEtAl2012, KabukcuEtAl2021, UlasEtAl2024}. Comparative
genetics also points to the Northern Mesopotamia region as the centre of
diversity of many crops \citep[e.g.][]{HaasEtAl2019}. But since these
studies are based on modern genomes, if the wild range of these plants
has, as our modelling suggests, shifted since the Early Holocene, the
apparent centre of diversity may have shifted with them. Our
reconstructions are consistent with the late arrival of intensive
plant-based foraging economies in this region (cf.~the Natufian of the
Levant), and more broadly there need not be a link between the core
\emph{wild} range of a plant and the core area of its domestication. A
scenario in which cultivation emerged at the edges of the ranges of
valuable plant resources---as a means of extend their natural niche---is
also plausible.

The near-absence of the Zagros in any predicted ranges is also
surprising, given mounting evidence that domestication took place just
as early in the eastern \emph{Mashriq} as it did in the west
\citep{Braidwood, GanjDarehGoats, HillyFlanksStuff}. We consider that
the most likely explanation for this is that our flora does not include
the species that were most important to plant subsistence in the east.
Archaeobotanical data on Neolithic sites in the Zagros is limited
(compared to the Levant in particular) due to a hiatus in field research
there from the 1980s to early 2010s \citep{cite}. Recent research
\citep{Amaia} indicates \ldots{}

Cyprus and the Aegean are not conventionally considered part of the
primary zone of domestication but rather amongst the first regions that
acquired agriculture from West Asia. Our analysis complicates this
picture, as it indicates that the wild ranges of many crop progenitors
included these regions. Early examples of several domesticates are
recorded at sites on Cyprus, Western Anatolia and Greece
\citep{ArranzOtaeguiRoe2023}, and the Aegean region was probably
connected to West Asia by a land bridge via Anatolia until the Early
Holocene \citep{AksuHiscott2022}. Were these area part of the same
broader `interaction sphere' \citep{cite} that produce Neolithic
agriculture in West Asia?

@@

Almost all the cereal and legume crop progenitors we modelled are
predicted to have only been found in the Levant during the terminal
Pleistocene and Early Holocene (see appendix). Part of this may be do
with the fact that both our initial flora and training occurrence
dataset have a strong bias towards the southern Levant, but strikingly
the modelled \emph{current} ranges of these plants do tend to include
Anatolia and the Zagros, so this cannot be the only factor. One notable
exception is the wild ancestor of chickpea (\emph{Cicer reticulatum}),
which is predicted to have a distribution centered on Northern
Mesopotamia but encompassing much of the the `hilly flanks' (except the
southern Levant, Figure~\ref{fig-palaeodist-chickpea-rye}). Another is
rye (\emph{Secale cereale}), which is inferred to be primarily Western
Anatolian Figure~\ref{fig-palaeodist-chickpea-rye}. This is perhaps
relevant to rye's unusual domestication history, as a crop of West Asian
origin that was apparently not first cultivated until much later than
the `founder crops', in Europe \citep{SchreiberEtAl2021}.

@@

Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation
Figure~\ref{fig-palaeodist-barely-levant}. It also sees a marked decline
in the archaeobotanical record from the Early PPNA/Early PPNB (where it
was amongst the most common taxa) to the Late PPNB and Late Neolithic
\citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia atlantica})
shows similar trends, but it is less certain that this species was
cultivated in the Neolithic. Conversely, the various wild wheat species
native to West Asia show almost no response to Pleistocene/Holocene
climate change, even within the Levant, and in the archaeobotanical
record wheat displays the opposite trend to barley and pistachio --
becoming gradually more abundant through the course of the Neolithic and
dominant by its end. We hypothesise that climate-linked range shifts
were a factor driving these changes in the apparent economic important
of different crops: the decreasing availability of barley and pistachio
in the terminal Pleistocene may have prompted cultivation as a strategy
to retain access to them, though in the long run it made them less
attractive as staple crops compared to the more resilient wheat.
However, these dynamics are not seen in the majority of crop
progenitors.

@@

``All models are wrong'' \citep{cite}, but the performance of our models
on independent test datasets give confidence in its predictions, which
are plausible and consistent with broad-scale patterns in the
archaeological record.

\begin{itemize}
\tightlist
\item
  Results w.r.t. to specific archaeologically-attested species not to
  promising, but this doesn't mean the models are useless!
\item
  Discrepencies suggests several areas for further research and
  methodological development
\end{itemize}

Modelling a large number of species using machine learning, the
substantial occurrence datasets available for the present day, and a
hindcasting approach to past distributions also represents a significant
advance in the methodology of palaeoecological niche modelling. This
approach is enabled by the availability of high quality, global open
datasets in ecology \citep{GBIF2025, GBIFSecretariat2023}, earth science
\citep{SRTM}, and climatology
\citep[@][]{KargerEtAl2017, BrownEtAl2018}. Unfortunately in relation to
these fields open data in archaeology lags conspiciously behind. Though
we have benefitted from the relatively good state of publication of
archaeobotanical compilations in our region of study
\citep{ADEMNES, COMPAG, ORIGINS}, further development of open,
comprehensive and up-to-date `backbone' datasets on site locations and
chronologies is needed to advance archaeoecological modelling to the
same level.

>>>>

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 82 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. Modelled ranges do not
reliably predict the observed occurrence of specific species at
archaeological sites, but are aligned with more general trends in the
relative abundance of species.
\end{abstract}

@@

The Pleistocene--Holocene transition in West Asia marked a turning point
in global environmental history, as humans brought the first plants
under cultivation and began modifying surrounding ecosystems to support
their own subsistence. West Asia is part of the native range of a
remarkable number of domesticable plant species, including wild
relatives of wheat, barley, peas, lentils, and other crops of global
importance \citep{Diamond2002}. These species supported uniquely dense
and complex Late Epipalaeolithic (15--11.7 ka) societies based on
intensive foraging \citep{BarYosef1998, MaherEtAl2012} and eventually
plant management and pre-domestication cultivation \citep{Harris2007}.
As they were brought under domestication in the Pre-Pottery Neolithic
period (11.7--8.5 ka), the world's first agriculture was shaped by the
ecosystems from which it emerged and was embedded in.

Decades of research in archaeobotany and zooarchaeology have
reconstructed the subsistence economies of Late Epipalaeolithic and
Neolithic sites in great detail. Together with studies of charcoal,
pollen, soil, and a variety of palaeoclimate archives
\citep{JonesEtAl2019}, they also tell us much about the environments
surrounding these settlements. However, each of these sources of
evidence is subject to the wide variety of taphonomic and recovery
biases that are inherent in any direct record of the past. They are
also, by definition, records of the (human) environment at particular
times and places. Interpolating these snapshots to give a holistic
picture of the regional ecologies is not straightforward -- to date, it
has tended to rely on non-explicit, inductive modelling. The majority
are also filtered through human action, producing a mixed single that
makes it difficult to disentangle anthropic effects from the background
of environmental change in this period of rapid climatic alteration.

@@

Ecological niche modelling or species distribution modelling is widely
used by ecologists to predict the geographic range of a species based on
a set of environmental predictors \citep{FranklinMiller2009}.
Essentially, it involves combining records of where an organism has been
observed with environmental data (climate, topography, etc.) for those
locations to model the range of environmental values at which that
species -- its environmental niche. This model can then be used to
predict the range of the organism in question either in the same or a
different environment. \citet{TownsendPetersonSoberon2012} suggests
reserving the term `species distribution modelling' for when the method
is used to recover the verifiable range of a species in a real and
existing environment, and using `ecological niche modelling' as the
broader term covering hypothetical or predictive applications -- a
convention we follow here when referring to predictive or `hindcast'
models of past ranges. Within this overarching framework, ecological
niche modelling encompasses a wide range of applications and a variety
of potential environmental predictors, modelling approaches, and
methodologies, which we will not attempt to review here.

@@

\citet{FranklinEtAl2015} review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. In an
early application to West Asia, \citet{ConollyEtAl2012} used the
occurrence of wild and domestic \emph{Bos} remains at prehistoric
archaeological sites to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales
\citep[e.g.][]{deAndresHerreroEtAl2018, YaworskyEtAl2023} and, in a West
Asian context, of foraged plant resources in the landscape around the
Neolithic sites on the Konya Plain \citep{CollinsEtAl2018}. Modelling
the spread of crops has been another significant archaeological
application \citep[e.g.][]{KrzyzanskaEtAl2022, Krzyzanska2023}, though
not as yet applied to West Asia.

@@

The major practical limitation of the hindcasting approach is that it
relies on spatially explicit, high resolution palaeoenvironmental
surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most
applications, which is perhaps why only a minority of studies use it
\citep[cf.][]{KrzyzanskaEtAl2022, YaworskyEtAl2023}. In this study, we
are able to take advantage of the increasing availability of high
resolution, global palaeoclimate data derived from simulation
experiments with general circulation models of climate
\citep{BrownEtAl2018, BrownEtAl2020, KargerEtAl2023}.

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 97
taxa (Table~\ref{tbl-results-summary}) comprising the identifiable
species observed at at least two Late Epipalaeolithic/Pre-Pottery
Neolithic sites, according to our previous study of the regional
archaeobotanical data \citep{ArranzOtaeguiRoe2023}. This was based on
dataset collated from three previously published regional
archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
\citep{ORIGINS}, and COMPAG
\citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
on][]{ColledgeEtAl2004}; \citealp{ShennanConolly2007}}. We did not
attempt to distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

@@

@@

For training, test predictions, and archaeological predictions predictor
data was left in its native projection and resolution. For hindcast
palaeodistributions, it was transformed to common equal-area projection
and resolution of 5 km.

@@

@@

Our results for the Levant are consistent with the current understanding
of this region as developing early intensive foraging economies
\citep[the Natufian culture,][]{BarYosef1998} and as a centre of origin
of agriculture \citep{Zeder2011a}. Within the Levant, many species show
moderate reductions in ranges over the Pleistocene/Holocence boundary,
retreating from the Badia/transjordan region (e.g.~barley, fig.
Figure~\ref{fig-palaeodist-barely-levant}).

Loss of the Northern Mesopotamia--Anatolia region from the predicted
ranges of crop progenitors is interesting in light of the `golden
triangle' hypothesis
\citep{Lev-YadunEtAl2000, KozlowskiAurenche2005, AbboEtAl2010}, which
puts this region at the centre of the development of agriculture.
Multiple lines of archaeological evidence have emerged that point away
from this hypothesis and towards a more geographically diverse origin
\citep{Asouti2006, FullerEtAl2011, ArranzOtaeguiEtAl2016}, but it
remains the area with some of the earliest clear evidence of
domestication \citep{ZoharyEtAl2012, KabukcuEtAl2021, UlasEtAl2024}.
Comparative genetics also points to the Northern Mesopotamia region as
the centre of diversity of many crops \citep[e.g.][]{HaasEtAl2019}. But
since these studies are based on modern genomes, if the wild range of
these plants has, as our modelling suggests, shifted since the Early
Holocene, the apparent centre of diversity may have shifted with them.
Our reconstructions are consistent with the late arrival of intensive
plant-based foraging economies in this region (cf.~the Natufian of the
Levant), and more broadly there need not be a link between the core
\emph{wild} range of a plant and the core area of its domestication. A
scenario in which cultivation emerged at the edges of the ranges of
valuable plant resources---as a means of extend their natural niche---is
also plausible.

The near-absence of the Zagros in any predicted ranges is also
surprising, given mounting evidence that domestication took place just
as early in the eastern \emph{Mashriq} as it did in the west
\citep{Zeder2024}. We consider that the most likely explanation for this
is that our flora does not include the species that were most important
to plant subsistence in the east. Archaeobotanical data on Neolithic
sites in the Zagros is limited (compared to the Levant in particular)
due to a hiatus in field research there from the 1980s to early 2010s
\citep{Zeder2024}. Recent research
\citep{WhitlamEtAl2018, GonzalezCarreteroEtAl2023} indicates that plant
subsistence in this region was based on a distinct set of species,
compared to the Levant and Anatolia.

Cyprus and the Aegean are not conventionally considered part of the
primary zone of domestication but rather amongst the first regions that
acquired agriculture from West Asia. Our analysis complicates this
picture, as it indicates that the wild ranges of many crop progenitors
included these regions. Early examples of several domesticates are
recorded at sites on Cyprus, Western Anatolia and Greece
\citep{ArranzOtaeguiRoe2023}, and the Aegean region was probably
connected to West Asia by a land bridge via Anatolia until the Early
Holocene \citep{AksuHiscott2022}. Were these area part of the same
broader `interaction sphere' that produce Neolithic agriculture in West
Asia?

@@

Almost all the cereal and legume crop progenitors we modelled are
predicted to have only been found in the Levant during the terminal
Pleistocene and Early Holocene (see appendix). Part of this may be do
with the fact that both our initial flora and training occurrence
dataset have a strong bias towards the southern Levant, but strikingly
the modelled \emph{current} ranges of these plants do tend to include
Anatolia and the Zagros, so this cannot be the only factor. One notable
exception is the wild ancestor of chickpea (\emph{Cicer reticulatum}),
which is predicted to have a distribution centered on Northern
Mesopotamia but encompassing much of the the `hilly flanks' (except the
southern Levant, Figure~\ref{fig-palaeodist-chickpea-rye}). Another is
rye (\emph{Secale cereale}), which is inferred to be primarily Western
Anatolian (Figure~\ref{fig-palaeodist-chickpea-rye}). This is perhaps
relevant to rye's unusual domestication history, as a crop of West Asian
origin that was apparently not first cultivated until much later than
the `founder crops', in Europe \citep{SchreiberEtAl2021}.

@@

Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation
(Figure~\ref{fig-palaeodist-barely-levant}). It also sees a marked
decline in the archaeobotanical record from the Early PPNA/Early PPNB
(where it was amongst the most common taxa) to the Late PPNB and Late
Neolithic \citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia
atlantica}) shows similar trends, but it is less certain that this
species was cultivated in the Neolithic. Conversely, the various wild
wheat species native to West Asia show almost no response to
Pleistocene/Holocene climate change, even within the Levant, and in the
archaeobotanical record wheat displays the opposite trend to barley and
pistachio -- becoming gradually more abundant through the course of the
Neolithic and dominant by its end. We hypothesise that climate-linked
range shifts were a factor driving these changes in the apparent
economic important of different crops: the decreasing availability of
barley and pistachio in the terminal Pleistocene may have prompted
cultivation as a strategy to retain access to them, though in the long
run it made them less attractive as staple crops compared to the more
resilient wheat. However, these dynamics are not seen in the majority of
crop progenitors.

@@

``All models are wrong'' \citep{Box1976}, but the performance of our
models on independent test datasets give confidence in its predictions,
which are plausible and consistent with broad-scale patterns in the
archaeological record. Modelling a large number of species using machine
learning, the substantial occurrence datasets available for the present
day, and a hindcasting approach to past distributions also represents a
significant advance in the methodology of palaeoecological niche
modelling. This approach is enabled by the availability of high quality,
global open datasets in ecology \citep{GBIF2025, GBIFSecretariat2023},
earth science \citep{SRTM}, and climatology
\citep{KargerEtAl2017, BrownEtAl2018}. Unfortunately in relation to
these fields open data in archaeology lags conspiciously behind. Though
we have benefitted from the relatively long tradition of compiling
archaeobotanical data in our region of study
\citep{ColledgeEtAl2004, ShennanConolly2007, ADEMNES, LucasFuller2018, FullerEtAl2018, WallaceEtAl2018, ORIGINS},
further development of open, comprehensive and up-to-date `backbone'
datasets on site locations and chronologies is needed to advance
archaeoecological modelling to the same level.
